\stktokens{}: Enforcing Well-Bracketed Control Flow and Stack Encapsulation Using Linear Capabilities

### 2019-07-23 popl-2019-journal-v/proving-full-abstraction.tex
[POPLjv] Insert missing definitions in cofe section

  An element $x \in X$ is the \emph{limit} of the sequence $\seq{x}$ if
  \begin{align*}
    \forall k \in \nats, \exists j \in \nats, \forall n \geq j, x \nequal[k] x_n.
  \end{align*}
\end{definition}
% TODO write something here about why the different things (the meassure of the ofe and the guaranteed limits of the cofe) gives us in terms of finding a solution.
In order to move from one c.o.f.e.\ to another, we define functions, just like we would when working with sets.
Unlike functions between sets, functions between c.o.f.e.'s must preserve the c.o.f.e.\ structure which informally means that the measure of precision must be retained.
This specifically means that functions between c.o.f.e.'s must be \emph{non-expansive} which basically means that they preserve $n$-equalities.
Functions may not only retain equivalences but make it more precise.
Such a function is said to be \emph{contractive}.
\begin{definition}
  \label{def:nonexpansive-contractive-ofe}
  Let $\left(X,\left(\nequal[n]_X\right)_{n=0}^{\infty} \right)$ and $\left(Y,\left(\nequal[n]_Y\right)_{n=0}^{\infty} \right)$ be two ordered families of equivalences and $f$ a function from the set $X$ to the set $Y$.
  We 
  The function $f$ is
  \begin{description}
  \item[non-expansive] when for all $x, x' \in X$, and all $n \in \nats$,
\[
  x \nequal_X x' \implies f(x) \nequal_Y f(x')
\]
  \item[contractive] when for any $x, x' \in X$, and any $n \in \nats$,
\[
  x \nequal_X x' \implies f(x) \nequal[n+1]_Y f(x') 
  \qedhere
\]
  \end{description}
\end{definition}
\noindent The step-indexing provided by c.o.f.e.'s give enough structure to solve the recursive domain equation.
There are a number of standard c.o.f.e.\ constructions for countably infinite sets, products, sums, and functions.
A set can be lifted to c.o.f.e.\ by having the normal equality for all positive indices and the total relation for $\nequal[0]$.
The other constructions are used to create c.o.f.e.'s from other c.o.f.e.'s.
For instance the product construction is defined as follows:
\begin{definition}[Product c.o.f.e.]
\label{def:product-cofe}
\emph{delete me}
\todo[inline]{insert definition}
\end{definition}
The remaining standard constructions are defined similarly.
Another important construction is the \emph{later} ($\blater$) c.o.f.e.\
The later c.o.f.e.\ construction is based on one c.o.f.e.\ which it decreases the index of.
\begin{lemma}[$\blater$ c.o.f.e.]
  \label{def:later-cofe}
  \todo[inline]{replace with actual definition for c.o.f.e (not preoreded!)}.
  Given a c.o.f.e $\left(X, \left(\nequal_X\right)_{n=0}^{\infty}, \future_X \right)$ define
  \[
    \blater\left(X, \left(\nequal_X\right)_{n=0}^{\infty}, \future_X \right) = \left(X, \left(\nequal[n]_{\blater}\right)_{n=0}^{\infty}, \future_X \right)
  \]
  where
  \[
    x \nequal[0]_{\blater} x' \wedge x \nequal[n]_{\blater} x' \text{ iff } x \nequal[n+1]_X x'
  \]
\end{lemma}
The later c.o.f.e.\ is used to make sure that the index decreases before the self-reference which gives something to recurse on in order to find a solution for the domain equation.
If we only use the standard constructions to construct a self-referential equation and make sure to guard every negative occurrence of the self reference with the later c.o.f.e.\ construction, then the equation has a solution.
\todo[inline]{Insert the solution theorem here?}

Unfortunately, the standard c.o.f.e.\ constructions do not suffice for the worlds we want to define.
The worlds we sketched have additional structure in terms of a future world relation.
The future world relation is captured by adding a preorder to c.o.f.e.'s:
\begin{definition}[Preordered c.o.f.e.]
  A preordered c.o.f.e.\ is a c.o.f.e.\ equipped with a preorder on $X$, $\left(X, \left(\nequal\right)_{n=0}^{\infty}, \future \right)$. 
  \begin{itemize}
  \item The ordering preserves limits. That is, for Cauchy sequences $\{a_n\}_n$ and $\{b_n\}_n$ in $X$ if $\{a_n\}_n \future\{b_n\}_n$, then $\lim \{a_n\}_n \future \lim \{b_n\}_n$.
  \end{itemize}
\end{definition}
The preorder must respect the existing structure of the c.o.f.e.'s which means that it must preserve limits.
There are also canonical constructions for preordered c.o.f.e.'s that simply rely on the preorder of the underlying preordered c.o.f.e.'s.
We will need the standard construction for later, product, and sum later, so we define them here.
The reader may skip these definitions for now and refer back to them later.
\begin{definition}[$\blater$ preordered c.o.f.e.]
  \label{def:later-p-cofe}
  \todo[inline]{replace with actual definition for preordered c.o.f.e}.
  Given a preordered c.o.f.e $\left(X, \left(\nequal_X\right)_{n=0}^{\infty}, \future_X \right)$ define
  \[
    \blater\left(X, \left(\nequal_X\right)_{n=0}^{\infty}, \future_X \right) = \left(X, \left(\nequal[n]_{\blater}\right)_{n=0}^{\infty}, \future_X \right)
  \]
  where
  \[
    x \nequal[0]_{\blater} x' \wedge x \nequal[n]_{\blater} x' \text{ iff } x \nequal[n+1]_X x'
  \]
\end{definition}
We will also need the canonical preordered c.o.f.e.\ for sum.
\begin{definition}[Product preordered c.o.f.e.]
  \label{def:product-p-cofe}
  \todo[inline]{Add definition}

\end{definition}
\begin{definition}[Sum preordered c.o.f.e.]
  \label{def:sum-p-cofe}
  \todo[inline]{Add definition}

@@

In the following, we will define a number of preordered c.o.f.e.'s that can be combined to get the world equation that we want.
First, we define the c.o.f.e.\ that we need for the memory invariants.
Note that we do not need a preorder on this.
As a reminder, we sketched the memory invariants to as follows
\[
  \World \monfun \Rel{\MemSeg \times \MemSeg}
\]
We will need to define a c.o.f.e.\ for the relation and a c.o.f.e.\ for the monotone and non-expansive functions.
The world will be given to us as the solution to the recursive equation.
Generally, we can lift a set of relations to a c.o.f.e.\ by adding an index to it that is downwards closed.
This means that if a pair is in the relation for a specific index, then it is also in the relation for all smaller indices.
%% URel
\begin{definition}[Uniform relation]
  \label{uniform-rel-cofe}
  Given a set $R = X \times Y$, we define $\URel{R} \subseteq \powerset{\nats \times R}$ as
  \[
    \URel{R} = \{ A \in \powerset{\nats \times R} \mid \forall n \in \nats \ldotp \forall r \in R \ldotp \npair{r} \in A \implies \forall m \leq n \ldotp \npair[m]{r} \in A\}
  \]
  The uniform relation c.o.f.e.\ $\cofe{\URel{R}}$ has $\URel{R}$ as the underlying set and the following family of equivalences:
  \[
    A \nequal B \text{ iff } \erasen{A}{n} = \erasen{B}{n}
  \]
  where erasure is defined as
  \[
    \erasen{A}{n} \defeq \{(k,a) \in A \mid k < n\}
  \]
\end{definition}
The uniform relation construction uses the canonical family of equivalences which basically ignores everything from the step and up and requires the rest to be equal.
This means that everything is equal at step $0$ as everything is left out.

@@

\begin{theorem}
  \label{thm:rec-dom-eq-sol}
  There exists a c.o.f.e.\ $\Wor$ and preorder $\future$ such that $(\Wor,\future)$ is a preordered c.o.f.e., and there exists an isomorphism $\xi$ such that
  \[
    \xi : \Wor \cong \blater (\Worldh \times \Worlds \times \Worldfs)
  \]
  and for $\hat{W},\hat{W}' \in \Wor$
  \[
    \hat{W}' \future \hat{W} \text{ iff } \xi(\hat{W}') \future \xi(\hat{W})
  \]
  for $\Worlds$, $\Worldh$, and $\Worldfs$ defined as follows
\[
  \Worldh = \RegionName \parfun (\Regions + \Regionh)
\]
and
\[
  \Worlds = \RegionName \parfun (\Regions \times \Addr)
\]
and
\[
  \Worldfs = \RegionName \parfun \Regions
\]
where $\RegionName = \nats$. $\Regions$ and $\Regionh$ defined as follows
\begin{multline*}
  \Regionh = 
  \{\pure \} \times (\Wor \monnefun \URel{\MemSeg^2}) \times \\
  (\Seal \parfun \Wor \monnefun \URel{\SealableCaps^2})
\end{multline*}
and
\todo[inline]{look over relations and make it uniform whether it is $\MemSeg\times\MemSeg$ or $\MemSeg^2$}
\[
  \Regions = \left\{
  \begin{array}{l}
    \{\spatial, \spatialo\} \times (\Wor \monnefun \URel{\MemSeg^2}) +\\
    \{\revoked\}
  \end{array} \right.
\]
\end{theorem}
% TODO informally describe how solution to recursive domain equation is found (in terms of what we have to show, not the fixed-point theorem).
%% The category-theoretic solution of recursive metric-space equations, Birkedal et al?
%% Aleš note? Tutorial note?
The solution to the recursive domain equation is the preordered c.o.f.e.\ $\Wor$ in Theorem~\ref{thm:rec-dom-eq-sol}.
Note, that $\Wor$ is not equal to $\blater (\Worldh \times \Worlds \times \Worldfs)$; it is isomorphic.
This means that whenever we encounter $\Wor$, we have to apply $\xi$ and go under a later before we can actually use the world.
This makes it rather inconvenient to have $\Wor$ as the world, so instead we define the worlds as 
\[
  \World = \Worldh \times \Worlds \times \Worldfs
\]

>>>>

  An element $x \in X$ is the \emph{limit} of the sequence $\seq{x}$ if
  \begin{align*}
    \forall k \in \nats, \exists j \in \nats, \forall n \geq j, x \nequal[k] x_n.
  \end{align*}
\end{definition}
% TODO write something here about why the different things (the meassure of the ofe and the guaranteed limits of the cofe) gives us in terms of finding a solution.
In order to move from one c.o.f.e.\ to another, we define functions, just like we would when working with sets.
Unlike functions between sets, functions between c.o.f.e.'s must preserve the c.o.f.e.\ structure which informally means that the measure of precision must be retained.
This specifically means that functions between c.o.f.e.'s must be \emph{non-expansive} which basically means that they preserve $n$-equalities.
Functions may not only retain equivalences but make it more precise.
Such a function is said to be \emph{contractive}.
\begin{definition}
  \label{def:nonexpansive-contractive-ofe}
  Let $\left(X,\left(\nequal[n]_X\right)_{n=0}^{\infty} \right)$ and $\left(Y,\left(\nequal[n]_Y\right)_{n=0}^{\infty} \right)$ be two ordered families of equivalences and $f$ a function from the set $X$ to the set $Y$.
  We 
  The function $f$ is
  \begin{description}
  \item[non-expansive] when for all $x, x' \in X$, and all $n \in \nats$,
\[
  x \nequal_X x' \implies f(x) \nequal_Y f(x')
\]
  \item[contractive] when for any $x, x' \in X$, and any $n \in \nats$,
\[
  x \nequal_X x' \implies f(x) \nequal[n+1]_Y f(x') 
  \qedhere
\]
  \end{description}
\end{definition}
\noindent The step-indexing provided by c.o.f.e.'s give enough structure to solve the recursive domain equation.
There are a number of standard c.o.f.e.\ constructions for countably infinite sets, products, sums, and functions.
A set can be lifted to c.o.f.e.\ by having the normal equality for all positive indices and the total relation for $\nequal[0]$.
The other constructions are used to create c.o.f.e.'s from other c.o.f.e.'s.
For instance the product construction is defined as follows:
\begin{definition}[Product c.o.f.e.]
\label{def:product-cofe}
Given two c.o.f.e.'s $\left(X, \left(\nequal_X\right)_{n=0}^{\infty} \right)$ and $\left(Y, \left(\nequal_Y\right)_{n=0}^{\infty} \right)$ define the product c.o.f.e.\ as $\left(X \times Y, \left(\nequal \right)_{n=0}^{\infty} \right)$ where the equivalence family is defined as for $(x,y),(x',y') \in X \times Y$
\[
  (x,y) \nequal (x',y') \text{ iff } x \nequal x' \wedge y \nequal y'
\]
\end{definition}
The remaining standard constructions are defined similarly.
Another important construction is the \emph{later} ($\blater$) c.o.f.e.\
The later c.o.f.e.\ construction is based on one c.o.f.e.\ which it decreases the index of.
\begin{lemma}[$\blater$ c.o.f.e.]
  \label{def:later-cofe}
  Given a c.o.f.e $\left(X, \left(\nequal_X\right)_{n=0}^{\infty} \right)$ define the later c.o.f.e.\ as
  \[
    \blater\left(X, \left(\nequal_X\right)_{n=0}^{\infty} \right) = \left(X, \left(\nequal[n]_{\blater}\right)_{n=0}^{\infty} \right)
  \]
  where
  \[
    x \nequal_{\blater} x' \text{ iff } \left\{
      \begin{array}{l}
        n = 0 \vee{}\\
        x \nequal[n+1]_X x'
      \end{array}
    \right.
  \]
\end{lemma}
The later c.o.f.e.\ is used to make sure that the index decreases before the self-reference which gives something to recurse on in order to find a solution for the domain equation.
If we only use the standard constructions to construct a self-referential equation and make sure to guard every negative occurrence of the self reference with the later c.o.f.e.\ construction, then the equation has a solution.
\todo[inline]{Insert the solution theorem here?}

Unfortunately, the standard c.o.f.e.\ constructions do not suffice for the worlds we want to define.
The worlds we sketched have additional structure in terms of a future world relation.
The future world relation is captured by adding a preorder to c.o.f.e.'s:
\begin{definition}[Preordered c.o.f.e.]
  A preordered c.o.f.e.\ is a c.o.f.e.\ equipped with a preorder on $X$, $\left(X, \left(\nequal\right)_{n=0}^{\infty}, \future \right)$. 
  \begin{itemize}
  \item The ordering preserves limits. That is, for Cauchy sequences $\{a_n\}_n$ and $\{b_n\}_n$ in $X$ if $\{a_n\}_n \future\{b_n\}_n$, then $\lim \{a_n\}_n \future \lim \{b_n\}_n$.
  \end{itemize}
\end{definition}
The preorder must respect the existing structure of the c.o.f.e.'s which means that it must preserve limits.
There are also canonical constructions for preordered c.o.f.e.'s that simply rely on the preorder of the underlying preordered c.o.f.e.'s.
We will need the standard construction for later, product, and sum later, so we define them here.
The reader may skip these definitions for now and refer back to them later.
\begin{definition}[$\blater$ preordered c.o.f.e.]
  \label{def:later-p-cofe}
  Given a preordered c.o.f.e $\left(X, \left(\nequal_X\right)_{n=0}^{\infty}, \future_X \right)$ define
  \[
    \blater\left(X, \left(\nequal_X\right)_{n=0}^{\infty}, \future_X \right) = \left(X, \left(\nequal[n]_{\blater}\right)_{n=0}^{\infty}, \future_X \right)
  \]
  where
  \[
    x \nequal_{\blater} x' \text{ iff } \left\{
      \begin{array}{l}
        n = 0 \vee{} \\
        x \nequal
        [n+1]_X x'
      \end{array}
    \right.
  \]
\end{definition}
We will also need the canonical preordered c.o.f.e.\ for sum.
\begin{definition}[Product preordered c.o.f.e.]
  \label{def:product-p-cofe}
  Given two c.o.f.e.'s $\left(X,\left(\nequal[n]_X\right)_{n=0}^{\infty}, \future_X \right)$ and $\left(Y,\left(\nequal[n]_{Y}\right)_{n=0}^{\infty}, \future_y \right)$, define the product preordered c.o.f.e.\ as
  \[
  \left(X \times Y ,\left(\nequal[n]\right)_{n=0}^{\infty},\future \right)
  \]
  where the preorder $\future$ distributes to the underlying preorder, i.e.\
  \[
\text{for } (x,y),(x',y') \in X \times Y, \quad    (x',y') \future (x,y) \text{ iff } x' \future_X x \wedge y' \future_Y y
  \]
  and the family of equivalences distributes to the underlying families of equivalences, i.e.\
  \[
\text{for } (x,y),(x',y') \in X \times Y, \quad    (x,y) \nequal (x',y') \text{ iff } x \nequal_X x' \wedge y \nequal_Y y'
  \]
\end{definition}
\begin{definition}[Sum preordered c.o.f.e.]
  \label{def:sum-p-cofe}
  Given two c.o.f.e.'s $\left(X,\left(\nequal[n]_X\right)_{n=0}^{\infty}, \future_X \right)$ and $\left(Y,\left(\nequal[n]_Y\right)_{n=0}^{\infty}, \future_y \right)$, define the product preordered c.o.f.e.\ as
  \[
  \left(X + Y ,\left(\nequal[n]\right)_{n=0}^{\infty},\future \right)
  \]
  where the preorder $\future$ distributes to the underlying preorder, i.e.\ 
  \[
    \text{for } z,z' \in X + Y,\quad  z' \future z \text{ iff } \left\{
      \begin{array}{l}
        z,z' \in X \wedge z' \future_x z \vee{} \\
        z,z' \in Y \wedge z' \future_Y z
      \end{array}
    \right.
  \]
  and the family of equivalences distributes to the underlying families of equivalences, i.e.\
  \[
    \text{for } z,z' \in X + Y,\quad  z \nequal z' \text{ iff } \left\{
      \begin{array}{l}
        z,z' \in X \wedge z \nequal_x z' \vee{} \\
        z,z' \in Y \wedge z \nequal_Y z'
      \end{array}
    \right.
  \]
\end{definition}

@@

In the following, we will define a number of preordered c.o.f.e.'s that can be combined to get the world equation that we want.
First, we define the c.o.f.e.\ that we need for the memory invariants.
Note that we do not need a preorder on this.
As a reminder, we sketched the memory invariants to as follows
\[
  \World \monfun \Rel{\MemSeg \times \MemSeg}
\]
We will need to define a c.o.f.e.\ for the relation and a c.o.f.e.\ for the monotone and non-expansive functions.
The world will be given to us as the solution to the recursive equation.
Generally, we can lift a set of relations to a c.o.f.e.\ by adding an index to it that is downwards closed.
This means that if a pair is in the relation for a specific index, then it is also in the relation for all smaller indices.
%% URel
\begin{definition}[Uniform relation]
  \label{def:uniform-rel-cofe}
  Given a set $R = X \times Y$, we define $\URel{R} \subseteq \powerset{\nats \times R}$ as
  \[
    \URel{R} = \{ A \in \powerset{\nats \times R} \mid \forall n \in \nats \ldotp \forall r \in R \ldotp \npair{r} \in A \implies \forall m \leq n \ldotp \npair[m]{r} \in A\}
  \]
  The uniform relation c.o.f.e.\ $\cofe{\URel{R}}$ has $\URel{R}$ as the underlying set and the following family of equivalences:
  \[
    A \nequal B \text{ iff } \erasen{A}{n} = \erasen{B}{n}
  \]
  where erasure is defined as
  \[
    \erasen{A}{n} \defeq \{(k,a) \in A \mid k < n\}
  \]
\end{definition}
The uniform relation construction uses the canonical family of equivalences which basically ignores everything from the step and up and requires the rest to be equal.
This means that everything is equal at step $0$ as everything is left out.

@@

\begin{theorem}
  \label{thm:rec-dom-eq-sol}
  There exists a c.o.f.e.\ $\Wor$ and preorder $\future$ such that $(\Wor,\future)$ is a preordered c.o.f.e., and there exists an isomorphism $\xi$ such that
  \[
    \xi : \Wor \cong \blater (\Worldh \times \Worlds \times \Worldfs)
  \]
  and for $\hat{W},\hat{W}' \in \Wor$
  \[
    \hat{W}' \future \hat{W} \text{ iff } \xi(\hat{W}') \future \xi(\hat{W})
  \]
  for $\Worlds$, $\Worldh$, and $\Worldfs$ defined as follows
\[
  \Worldh = \RegionName \parfun (\Regions + \Regionh)
\]
and
\[
  \Worlds = \RegionName \parfun (\Regions \times \Addr)
\]
and
\[
  \Worldfs = \RegionName \parfun \Regions
\]
where $\RegionName = \nats$. $\Regions$ and $\Regionh$ defined as follows
\begin{multline*}
  \Regionh = 
  \{\pure \} \times (\Wor \monnefun \URel{\MemSeg\times\MemSeg}) \times \\
  (\Seal \parfun \Wor \monnefun \URel{\SealableCaps\times\SealableCaps})
\end{multline*}
and
\[
  \Regions = \left\{
  \begin{array}{l}
    \{\spatial, \spatialo\} \times (\Wor \monnefun \URel{\MemSeg\times\MemSeg}) +\\
    \{\revoked\}
  \end{array} \right.
\]
\end{theorem}
% TODO informally describe how solution to recursive domain equation is found (in terms of what we have to show, not the fixed-point theorem).
%% The category-theoretic solution of recursive metric-space equations, Birkedal et al?
%% Aleš note? Tutorial note?
The solution to the recursive domain equation is the preordered c.o.f.e.\ $\Wor$ in Theorem~\ref{thm:rec-dom-eq-sol}.
Note, that $\Wor$ is not equal to $\blater (\Worldh \times \Worlds \times \Worldfs)$; it is isomorphic.
This means that whenever we encounter $\Wor$, we have to apply $\xi$ and go under a later before we can actually use the world.
This makes it rather inconvenient to have $\Wor$ as the world, so instead we define the worlds as 
\[
  \World = \Worldh \times \Worlds \times \Worldfs
\]
